---
title: "FastQ in Bioconductor exercises"
author: "Tom Carroll"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,eval=T)
```


## Installing Bioconductor packages

In these exercises we will review some of the **ShortRead** packages functionality with handling fastQ files.

The example data can be found in the data directory 

**Data/sampled_ENCFF000CXH.fastq.gz**

We will also use  the full FastQ file from this dataset which you can download from [here.](https://www.encodeproject.org/files/ENCFF000CXH/@@download/ENCFF000CXH.fastq.gz).

1. Read in the **sampled_ENCFF000CXH.fastq.gz** file.

```{r installdeBSded,include=FALSE}
suppressPackageStartupMessages(library(ShortRead))
```

```{r installBSded}
library(ShortRead)
fastqFile <- readFastq("../Data/sampled_ENCFF000CXH.fastq.gz")
```

2. Plot the occurrence of A, G, T, C and N in the read with ID **"42JV5AAXX_HWI-EAS229_1:6:1:633:2029"** using ggplot2

```{r installBS2w}
fastqRead <- fastqFile[id(fastqFile) == "42JV5AAXX_HWI-EAS229_1:6:1:633:2029"]
fastqReadSeq <- sread(fastqRead)
alpFreq <- alphabetFrequency(fastqReadSeq)
alpFreq5 <- alpFreq[,c("A","C","G","T","N")]
toPlot <- data.frame(bases=names(alpFreq5),Freq=alpFreq5)
library(ggplot2)
ggplot(toPlot,aes(x=bases,y=Freq,fill=bases))+geom_bar(stat="identity")+theme_bw()
```

3. Create a boxplot of quality scores over cycles for reads using base graphics.

```{r installBSww2}
allQuals <- quality(fastqFile)
forBox <- as(allQuals,"matrix")
colnames(forBox) <- paste0("Cycle",1:ncol(forBox))
boxplot(forBox)
```

4. Plot the frequency of A, G, C, T and N bases over cycles using ggplot2/

```{r installBS2de}
alpByCyle <- alphabetByCycle(sread(fastqFile))
alpByCyleFilt <-  alpByCyle[c("A","G","C","T","N"),]
AbyCycFrame <- data.frame(Base="A",Freq=alpByCyleFilt["A",],Cycle=1:76)
CbyCycFrame <- data.frame(Base="C",Freq=alpByCyleFilt["C",],Cycle=1:76)
TbyCycFrame <- data.frame(Base="T",Freq=alpByCyleFilt["T",],Cycle=1:76)
GbyCycFrame <- data.frame(Base="G",Freq=alpByCyleFilt["G",],Cycle=1:76)
NbyCycFrame <- data.frame(Base="N",Freq=alpByCyleFilt["N",],Cycle=1:76)
myFrame <- rbind(AbyCycFrame,CbyCycFrame,TbyCycFrame,GbyCycFrame,NbyCycFrame)
ggplot(myFrame,aes(x=Cycle,y=Freq,colour=Base))+geom_line()+theme_bw()

```

5. Read in a random 100000 bases from the downloaded **ENCFF000CXH.fastq.gz** file

```{r installBS2ecdde,cache=TRUE}
mySampledFile <- FastqSampler("~/Projects/Results/RNAseqPipeTest/FirstTest/FQs/ENCFF000CXH.fastq.gz",n=100000)
myFastQ <- yield(mySampledFile)
```

6. Count the number of duplicate sequences and filter file to only the non-duplicated sequences


```{r instjnknallBS2drre,dependson="installBS2ecdde"}
dupLogical <- srduplicated(myFastQ)
numberOfDups <- table(dupLogical)
nonDups <- myFastQ[dupLogical]
```

7. Trim the resulting non-duplicated reads to remove reads when a succession of 10 bases fall below quality of 5. 

```{r instaljnkklBS2drre,dependson="instjnknallBS2drre"}
trimmedNonDups <- trimTails(nonDups,10,"5")
```

8. Plot a histogram of the resulting read lengths using base graphics

```{r installBSi2drre,dependson="instaljnkklBS2drre"}
hist(width(trimmedNonDups),xlab="Trimmed Length",main="Histogram of trimmed lengths")
```

9.Filter out reads less than 32 bases and write to file
```{r installswswBS2kjdrre,dependson="installBSi2drre"}

unlink("filteredFastq.fq.gz")
```

```{r installBS2kjdrre,dependson="installswswBS2kjdrre"}
filteredFastq <- trimmedNonDups[width(trimmedNonDups) >= 32]
writeFastq(filteredFastq,file="filteredFastq.fq.gz")
```



